---
title: "A Study on A Home Team's Performance Relative to the Importance of Their Game"
author: "Jordy Chaves"
description: "How Barclay's Premier League clubs' forms in the 2016-2017 season were influenced by the importance of their games."
output:
  distill::distill_article:
    self_contained: false
---

# Introduction

  As a passionate soccer aficionado, I always hear pundits discussing the importance of games to a certain team. In any given game, there is a certain team that could use a win more than the other and perhaps there is an extra push from the players to get a result. Especially in league settings when the motivation is to qualify for continental competitions, avoid relegation, and even get more money because of their league position. So my primary research question became does the importance of a soccer game influence whether or not a home team wins a game? In this study, I want to observe the relationship between a match's importance to a home team and the result of that match for the team. My hypothesis is that as a match increases in importance, a home team will be more likely to get a positive result from their game. The null hypothesis would be that there is no relationship and that the variables are independent of each other. A positive result in this study is defined as a win or draw since wins are the results that yield the most points for a team (3 points). A negative result is defined as a loss or a draw (0 points and 1 point respectively) because the team failed to capitalize on all the points for the offering. 
  
The study is important in order to try and understand how best to predict a sports game. There are many variables that go into predicting the result of a soccer match, but if we can isolate the importance of a game as the explanatory variable, then we can form more accurate predictions. From the study, one could even start attempting to see these trends in other seasons and leagues to see if it remains valid.
  
# Data

The data I am using for this study comes from the `spi.matches.csv` dataset from FiveThirtyEight. It includes club soccer predictions and global club soccer rankings for 40 leagues. The names and descriptions of the variables included in this dataset are shown below. For this study, it is important to preface that the SPI (Soccer Power Index) ratings of the teams are generated by ESPN's rating system and is designed to best represent a team's current skill level. It is described as forward-looking and predictive. Additionally, this is an observational study as opposed to a randomized control trial. The outcome (result of the game) is not randomly assigned by chance. We are observing past results that were a result of many different variables that go into a soccer game. This observational study can help us generalize those performances for all teams in the Barclay's Premier League.

 My sample for this study is comprised of the 20 Barclay's Premier League teams in the 2016-2017 season. My unit of analysis will be a home team (for each individual game). The explanatory variable of interest is how important a game is to a home team (importance1 measures the importance of a home team). The explanatory variable is measured on a scale of 0-100, 0 meaning not important and 100 meaning the most important. My outcome variable is the results of a game for a home team. If I observe a positive linear relationship between the importance of a match to a home team and the result, then this will provide support for my hypothesis. If there is a negative relationship, then this would provide support against my hypothesis as this would mean that the more important the game, the less a home team wins.

Name                 Description
-------------------- -------------------------------------------------------------------------------------
`season`              The season during which the match was played. 
`date`                The date of the match (YYYY-MM-DD).
`league_id`           A unique identifier for the league this match was played in.
`league`             Name of the league this match was played in.      
`team1`              The home team's name.
`team2`              The away team's name.
`spi1`               The home team's overall SPI rating before the match.
`spi2`               The away team's overall SPI rating before the match.
`prob1`              The probability of the home team winning the match.
`prob2`              The probability of the away team winning the match.
`probtie`            The probability of match ending in a draw. (if applicable)
`proj_score1`        The number of goals we expected the home. team to score
`proj_score2`        The number of goals we expected the away team to score
`importance1`        The importance of the match for the home. team (0-100).
`importance2`        The importance of the match for the away. team (0-100).
`score1`             The number of goals scored by the home. team.
`score2`             The number of goals scored by the away. team.
`xg1`                The number of expected goals created by the home team.
`xg2`                The number of expected goals created by the away team.
`nsxg1`              The number of non-shot expected goals created by the home team.
`nsxg2`              The number of non-shot expected goals created by the away team.
`adj_score1`          The number of goals scored by the home team, adjusted for game state.
`adj_score2`          The number of goals scored by the away team, adjusted for game state.

## Loading the Data

```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
Data <- read_csv("Data/spi_matches.csv")
Data
```

## Subsetting the Data to Focus on the 2016 Barclay's Premier League Season and Creating New Variables

Because the scope of this study is on the 2016-2017 English Premier League, the data I am using needed to be subset. To do that, I set the variable `league` to "Barclay's Premier League" and `season` to "2016". Additionally, because this study is focused on the results of a home team, new variables had to be added to the data to assess the relationship between game importance and the results of those games. The new variables that I created in the study were `result` and `result1`. The former is just a categorical variable that assigns "Home win," "Home loss," and "Draw" for the corresponding scoreline of the match. `result1` is a binary variable that assigns `1` to games where the home team won and `0` to games where the home team either lost or tied. This was done by taking the `score1` and `score2` variables and using logical statements to represent wins, draws, and losses.


```{r}
Data <- Data |>
  drop_na()

pl_table_2016 <- Data |>
  filter(league == "Barclays Premier League", season == "2016") |>
  mutate(result = case_when(score1 > score2 ~ " Home Win",
                            score1 < score2 ~ "Home Loss",
                            score2 == score1 ~ "Draw"),
         result1 = case_when(score1 > score2 ~ 1,
                            score1 < score2 ~ 0,
                            score2 == score1 ~ 0)) |>
  group_by(importance1)
pl_table_2016
```  

## Histogram of Games and Their Importance

```{r}
pl_importance <- ggplot(data = pl_table_2016, mapping = aes(x = importance1)) +
    geom_histogram(fill = "deepskyblue3") +
    labs(x = "Degree of Importance of a Game to the Home Team", y = "Count", title = "Count of Games and How Important They Are") +
  theme_bw() +
  geom_vline(aes(xintercept = mean(pl_table_2016$importance1)), color = "red2")
  
pl_importance
```

This histogram visualizes the count of the importance of games. In the data, there are certainly more games that are less important than important. Most of the data is skewed to the left of the histogram. The vertical line shows the mean of the importance of games to a home team, which is about 37. So an average game's importance was about 37 points out of 100.

## Bar Plot of the Home Team's Results 

```{r}
count_barplot <- ggplot(data = pl_table_2016, mapping = aes(x = result, fill = result)) +
  geom_bar() +
  labs(x = "Result of Game", y = "Count", title = "Number of Wins, Draws, and Losses For a Home Team") +
  theme_bw()
count_barplot
```

This bar graph breaks down the dependent variable in this project: the result of games for a home team. Out of the 344 total games observed in the data, there are 166 home wins, 76 draws, and 102 home losses. In terms of negative and positive results, there are more negative results than positive results. The question becomes, in the absence of the actual scoreline, how much influence does the importance of a game have on the outcome? 


## Confounders

With the experiment, I wanted to account for potential confounding variables that would most influence the results of the games and the importance of the game. Two confounding variables I chose to analyze were the projected score for the home team and the home team's Soccer Power Index (SPI) rating before the game. The projected score could influence the result as a smaller projected score for the home team could lead to a negative result. The projected score may also be a reflection of the importance of a game as a higher projected score may mean a higher importance as teams have that pressure to achieve what is projected for them. SPI rating can make influence the result as a lower rated team may falter to higher rated team and the SPI rating might reflect the importance of the match. Facing similarly rated or worse-rated opposition may make it more imperative for teams to get a positive result. Facing higher opposition would mean a game is less important because they are not as likely to win. 

# Results

## Scatter Plot and Regression of Main Dependent Variable: 

My primary independent variable of interest was the `importance1` (Importance of a Game to a Home Team) variable. The scatter plot and linear regression below details the relationship between `result1` and `importance1`.

```{r}
pl_importance_results <- ggplot(data = pl_table_2016, mapping = aes(x = importance1, y = result1)) +
  geom_point(alpha = 0.3, color = "navy") +
  theme_bw() +
  geom_smooth(method = "lm", color = "dodgerblue2", fill = "skyblue") +
  labs(x = "Importance of the game to a Home Team", y = "Result of the Game (1 = win and 0 = loss/draw)", title = "Results of Games in Relation to the Importance of the Game to a Home Team")
pl_importance_results

imp_reg <- lm(result1 ~ importance1, data = pl_table_2016)
imp_names1 <- c("(Intercept)" = "Intercept", "importance1" = "Importance of Game to Home Team")

modelsummary::modelsummary(imp_reg, statistic = "p.value", 
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = imp_names1)
```

The main plot of analysis is the result of home teams' games in relation to the game's importance. The points on the scatter plot are on y = 0 and y = 1, because the result variable is a binary. 0 meaning a negative result (loss or a draw) and 1 meaning a positive result (win). The plot shows a higher congregation of negative results when the importance of the game is lower and a higher congregation of positive results when the importance is greater. As the unit of importance increases, we can see that there is a higher possibility of the home team winning the game. The regression line shows the positive association between the importance of games and its results for the home team. From this plot, the hypothesis can be substantiated, but this is done without adjusting for any confounding variables that might be influencing the results of a game. 

The linear regression expands on the scatter plot and provides concrete data. The table above shows that the intercept of the line is 0.336. This means that when the importance of a game to the home team is 0, they have a 33.6% chance of getting a win. The coefficient of `importance1` is 0.004 which means that for every point increase of the importance score, the home team increases their chances of winning by 0.4%. So if the importance of a game is scored as 100, then a team increases their chance of winning by 40%. The p-value of this regression is less than 0.001. From this initial observation of the main outcome of interest, there is great statistical significance. However, given that this is an observational study, I cannot interpret the relationship causally. There does seem to be a correlation between the importance and the result of the game, but we cannot confidently say that the importance of a game causes a team to get a win. 

## Multiple Regression Output

```{r}
library(texreg)
fit1 <- lm(result1 ~ importance1 + importance2 + xg1 + xg2 + spi1 + spi2 + proj_score1 + proj_score2, data = pl_table_2016)



fit2 <- lm(result1 ~ importance1, data = pl_table_2016)

library(broom)
fit1_table <- tidy(fit1)
fit1_table

pl_var_labels <- c("(Intercept)" = "Intercept",
                  "importance1" = "Match Importance to Home Team",
                  "importance2" = "Match Importance to Away Team",
                  "xg1" = "Home Team Expected Goals",
                  "xg2" = "Away Team Expected Goals",
                  "spi1" = "Home Team's Rating Before the Match",
                  "spi2" = "Away Team's Rating Before the Match",
                   "proj_score1" = "Projected Score for the Home Team", 
                  "proj_score2" = "Projected Score for the Away Team"
)

modelsummary::modelsummary(fit1, statistic = c("s.e. = {std.error}", "p = {p.value}"), coef_map = pl_var_labels, gof_map = c("nobs", "r.squared", "adj.r.squared"), fmt = 3)
```

In the table above, the intercept is 0.583. This implied when all variables are 0 (in this case, the result is not important to home or away teams, expected goals for home and away teams are 0, SPI rating for home and away teams are 0, and the projected score for the home and away team is 0) then the probability of winning a home team winning a game is 58.3%. This specific intercept projects that a home team has a slight advantage over their opponents. Within the variables, the "projected score for the away team" variable has the largest effect with 0.499,meaning that the probability of a home team winning increases 49.9% per unit increase in an away team's projected score, holding all else equal. Finally, the "match importance to home team" variable has a coefficient of 0.002, which implies that per unit increase in match importance, the probability of the home team winning increases by 0.2%, holding all other variables equal. This number is consistent with the positive relationship observed in the scatter plot, it is lower than the 0.04% increase in the previous scatter plot and linear regression table. This implies that some of the correlation might be due to one of the variables in this regression. The p-value of the "match importance to the home team" variable is 0.036 which is below the alpha level of 0.1 but above the alpha level 0.05, which implies that we cannot conclude that there is statistical significance. We cannot interpret the results of this experiment causally given that the study has a cross-sectional design.


## Confounding Variables

Taking a closer look at the confounding variables I set out to analyze, `proj_score1`and `spi1`, I included them as covariates in the multiple regression table. The "projected score for the home team" variable had a coefficient of -0.323 meaning that for an increase of 1 unit in the projected score, the chance of a home team winning actually decreases by 32.3%, holding all others equal. The p-value of the variable is 0.106 which is above the alpha level of 0.1 which shows that this variable is not statistically significant. The "Home Team's Rating Before the Match" variable had a coefficient of 0.022 which means that for a 1-unit increase in the team's rating, the chances of the home team winning increased by 2.2%, holding all others equal. The p-value of the variable is 0.024 which is below the alpha level of 0.05, making this statistically significant. Based off of this regression, a home team's rating is a better metric for predicting the result of a game.

# Conclusion

In conclusion, there is an association between the importance of a game to a home team and the result in the 2016-2017 Barclay's Premier League season. In my regressions and scatter plot, the coefficients for the variable in question were not equal to zero, but they were not the best predictor of a win. The best predictor for a home win was the projected score for the away team. Additionally, the confounder of a team's SPI rating did prove to be a better predictor for a result within the regression. The coefficient of this confounder was greater than that of importance and the p-value suggested that it is statistically significant. Though these results show an association, it does not substantiate my hypothesis that a greater importance of a game would cause a home team to win. Therefore, these results support a rejection of the null hypothesis, but we cannot assume causation. There are some limitations to this study. For one, the data has a small sample size even if it has strong external validity. Another limitation to the study is the . With more time, I would have liked to conduct this study on other leagues separately. Perhaps the results would be different in countries such as Germany, France, Mexico, Spain, etc. Perhaps broadening the scope of the study can increase its external validity. Finally, it would be prudent to factor in a team's play style as a potential covariate as that could potentially help predict the result of a game.
